Executive Constraints and Civil Conflict Onset

### 2019-09-12 paper.Rmd
Update stan syntax highlighting xml path
WIP: paper updates

---
title: Executive Constraints and Civil Conflict
author: J.L.A. Krusell
output: pdf_document
---

\[ x_{j}^{obs} \sim \mathsf{Normal}(x_{j}^\ast, \sigma_{j}^{obs}) \]
\[ x_{ji}^\ast \sim \mathsf{Normal}(\gamma_j + \lambda_j \theta_i, \psi_j) \]
\[ \theta \sim \mathsf{Normal}(0, 1) \]

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: \today
documentclass: scrartcl
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
output:
    pdf_document:
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
---

```{r include = F, results = "hide"}
knitr::opts_chunk$set(echo = F, eval = F)

library(dplyr)
library(precrec)
library(scales)
library(thesis.utils)

```

## Research Design

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian multilevel logistic regression. In what follows, I first
describe the independent and dependentvariables, including the
operationalizations and data sources, and then delve into the
regression model before finally presenting the results.

### Civil Conflict

The dependent variable, *civil conflict onset*, was taken from the
UCDP/PRIO Armed Conflict Dataset v19.1 (Cite) covering 1946-2018
where intra-state conflict is defined as a violent contest over a
government or territory between a state and non-state actor with a
minimum threshold of 25 battle deaths within a single country-year.
The data was merged based on country location for conflicts labeled as
internal or internationalized internal --- *i.e.* where
`type_of_conflict` was either 3 or 4. Multiple conflicts within the
same years were aggregated to form a binary variable coded as 1 to
denote the onset of at least one new civil conflict and 0 for all
other years, including years of ongoing conflict.^[Extrasystemic
conflicts, *i.e.* imperial or colonial conflicts, were excluded
largely due to lack of data. Ideally, such conflicts would be merged
based on the territory location of the conflict, however, for
colonies, for example, there is very little easily available economic
data and it would be undesirable to use the data for the colonizing
power due to the often wide discrepancies in development within
colonial empires. TODO: maybe a blurb on theory?]

The primary onset dates were determined according to UCDP/PRIO's
episodic definition where a new onset is coded for the same
incompatibility when there is at least a one year pause in the
fighting or drop below the battle death threshold. The primary
rationale is that civil conflicts often go through several distinct
phases of fighting which can make it difficult to accurately code the
end of a conflict. The episodic definition avoids potentially
conflating briefs lulls involving the same rebel groups and leaders
versus the re-emergence of previously long dormant
conflicts. Furthermore, in both scenarios, the theory linking
executive constraints to the incentives for political violence should
still hold (TODO: possible expand here).

That said, because of the potential sensitivity of results to changing
conflict definitions (cite Sambanis), all analyses were also run using
only unique onsets.^[Likewise, appendix ? presents results
using a higher threshold for battle deaths. The end results stay the
same TODO]

### Executive Constraints

(TODO: Probably cite Sklar and/or O'Donnell here for more theoretical
justification of operationalization)

To measure formal executive constraints I use data from the Varieties
of Democracy (V-Dem) dataset v9.0 (cite V-Dem). V-Dem is a
cross-national time series dataset providing multidimensional,
disaggregated data on democracy and political institutions for over
200 countries with an overall year coverage from 1789 until 2018. The
principal data is drawn from expert coded surveys involving a global
network of over 3000 coders, mostly academics as well as professionals
working in policy related fields (cite Methodology doc). Individual
ratings are aggregated into latent variable estimates using
hierarchical ordinal item response theory (IRT) models taking into
account rater reliability and varying rater thresholds (cite Pemstein
et al). ^[Thresholds relate to how the continuous latent trait for a
given variable is mapped on to the ordinal answer categories presented
on the survey. Due to differential item functioning, experts may
perceive the magnitude of difference between answer items
differently. For example, for the variable *v2elvotbuy*, election vote
buying, experts may vary in how they interpret the difference between
a 2 (Restricted Vote Buying) and a 3 (Almost No Vote Buying). These
two concepts, rater reliability and varying rater thresholds,
correspond respectively to the *discrimination* and *difficulty*
parameters in classical IRT models.] The result is over 220 lower
level indicators that are also combined through various aggregation
techniques to produce over 80 mid/high-level indices. Compared to
datasets such as Polity IV (cite), which has traditionally been
used in most analyses of civil conflict, V-Dem provides much more fine
grained, nuanced data, and most importantly, includes measures of
uncertainty propagated from the IRT models. This latter point is
crucial since ignoring measurement error, especially in models with
noisy data and small effect sizes as is often the case in conflict
studies, can bias estimates and lead to overconfident results (cite
Treier & Jackman; Kyle).

For the purposes of this paper, I re-estimate Lührmann *et al*'s
horizontal accountability index (*v2x_horacc*) in V-Dem with the
addition of several variables (cite). The *v2x_horacc* index aims to
capture the extent to which the executive is *de facto* constrained by
other government institutions --- more specifically, the legislature,
judiciary, constitution, and independent investigative bodies (cite pg
61) --- rather than the *de jure* powers of any one institutions
vis-à-vis the executive. It is estimated as a latent variable using a
structural equation model, thereby avoiding the arbitrary weighting of
linear additive indices.^[A theoretical argument could be made that
the oversight offered by each institution is a necessary condition and
that any index should therefore be additive with equal
weighting. However, this would be a normative measure capturing the
degree to which a country-year fulfills an ideal concept, rather than
measuring how constrained an executive actually is in practice.]

For the two principal institutions, the legislature and the judiciary,
to serve as effective constraints there needs to be both a *capacity*
and *willingness* to oversee the executive. To better capture the
former dimension, I include two additional legislative variables. In
addition to *v2lgqstexp*, the extent to which the legislature
questions executive branch officials, and *v2lginvstp*, the extent to
which the legislature investigates in practice, I add *v2lgfunds*, the
extent to which the legislature controls its own resources, and
*v2lgoppart*, the extent to which opposition parties are able to
exercise oversight. The remaining manifest variables are the
same. Namely, for the judicial component: *v2juhcind*, high court
independence; *v2juncind*, lower court independence; *v2juhccomp*, the
degree to which the government complies with high court decisions;
*v2jucomp*, the degree to which the government complies with lower
court decisions.

v2exrescon
v2lgotovst

How measured? FA model, include formula

Measurement error (cite?)

Special case for legislative variables

Imputation for v2lgotovst

DAG

\[ x_{jt}^\ast \sim \mathsf{Normal}(\lambda_j \theta_t, \psi_t) \]
\[ x_{jt}^{obs} \sim \mathsf{Normal}(x_{jt}^\ast, \tau_j^{obs}) \]

### Controls

- state capacity (GDP per capita)

- gdp growth

- population density

- mean elevation

- relevant group counts

- neighbouring conflict

- ongoing conflict

- peace years

- varying intercepts

What's not included (political instability) and why

### Model

I model civil conflict onset as a bayesian multilevel regression where
onset is a function of a set of predictors, including executive
constraints and an array of control variables, together with several
grouping factors included as partially pooled varying intercepts. The
units of analysis are all independent country-years from 1946 until
2016, as determined by data availability, with *n* countries indexed
by $i = 1 \dotsc n$ and *m* years indexed by $t = 1 \dotsc m$. The
dependent variable, conflict onset, is a Bernoulli distributed binary
variable where 1 denotes the onset of at least one new civil conflict
in a given country-year. The chosen link function transforming the
linear regression output into the probability of conflict onset,
$\pi_{it}$, is the inverse logistic such that $\pi_{it} \in [0, 1]$.
^[Technically, the inverse logistic transforms values to the open
interval, $(0, 1)$, but because of overflow/underflow in computer
calculations due to floating point math the endpoints can be
observed.]

The full regression specification is then as follows:

\[ y_{it} \sim \mathsf{Bernoulli}(\pi_{it}) \]
\[ \pi_{it} = \mathsf{logit^{-1}}(\alpha + X_{it-1} \beta + \sum_{k=1}^K \delta^k_{j_{it}}) \]

where **X** is a single row matrix of predictors, $\beta$ a vector of
regression slopes, and $\delta^k$ the set of varying intercepts for
*K* grouping factors indexed by $j_k$. To account for time
endogeneity, all predictors are lagged by one year. They are also
standardized by subtracting the mean and dividing by the standard
deviation for each variable to aid in computation as well as
comparison of the final $\beta$ estimates (cite Gelman maybe?).

I place weakly informative priors on all parameters to allow the
posterior estimates to be largely determined by the observed data
while avoiding the more pathological values and hard constraints that
can arise from uninformative priors (cite Stan). For the slopes and
population intercept, given that the input data has been standardized,
this translates into

\[ \alpha \sim \mathsf{Normal}(0, 5) \]
\[ \beta \sim \mathsf{Normal}(0, 2.5) \]

Finally, for the varying intercepts modeling group-specific
variability, I use hierarchical priors that provide a regularizing
effect against overfitting (cite Gelman and Hill - Multilevel
Regression).

\[ \delta^k \sim \mathsf{Normal}(0, \sigma^k) \]
\[ \sigma^k \sim \mathsf{HalfCauchy}(0, 1) \]

## Analysis

Estimated using HMC by stan (footnote number of iterations, burnin, and convergence)

```{r}
beta <- readRDS("posteriors/summary/beta.rds")
plot_pars(beta)
```

```{r}
input.df <- readRDS("posteriors/input_data.rds")
p_hat <- readRDS("posteriors/summary/predicted_probs.rds")

plot_dens(p_hat, groups = input.df$lepisode_onset)
```

```{r}
p <- apply(p_hat, 2, median)
x <- evalmod(scores = p, labels = input.df$lepisode_onset)
plot(x, curvetype = c("ROC", "PRC"))
```

```{r eval = F}
# Interaction plot
intercept <- readRDS("posteriors/summary/intercept.rds")
theta <- readRDS("posteriors/summary/theta.rds") %>% apply(2, median) %>% sort

state <- quantile(input.df$cgdppc, probs = c(0.2, .5, 0.8))

f <- function(x, y) {
    with(input.df, intercept + beta[, 1] * x + beta[, 2] * y +
                   beta[, 3] * x * y + beta[, 4] * mean(peace_yrs) +
                   beta[, 5] * mean(gdpgro) + beta[, 6] * mean(pop_density)  +
                   beta[, 7] * mean(meanelev) + beta[, 8] * mean(rlvt_groups_count) +
                   beta[, 9] * 1 + beta[, 10] * 0)
}

par(mfrow = c(1, 3))
for (y in state) {
    print(y)
    p <- boot::inv.logit(sapply(theta, f, y)) * 100

    m <- apply(p, 2, median)
    pi <- apply(p, 2, quantile, probs = c(0.2, 0.8)) %>% t

    plot(theta, m, type = "l", ylim = c(min(pi[, 1]), max(pi[, 2])),
         xlab = "Executive Constraints", ylab = "Predicted Probability")
    lines(theta, pi[, 1], lty = 2)
    lines(theta, pi[, 2], lty = 2)
}

```

\newpage

# Appendix

## Stan Code

```{stan, code = readLines("stan/model.stan"), eval = FALSE, output.var = "x", echo = T}
```
